---
title: "Number of photos and pixels EDA and Models"
format: html
editor: visual
---

```{r}
#| output: false
#| code-fold: true
library(tidyverse)
tidy_data <- read_csv("../data/tidy_data/tidy_data_verifiable.csv")
tidy_data <- tidy_data %>% 
        select("id", "uri", "quality_grade", "created_at", "taxon_id", "n_photos",
               "microscopy_project","days_since_upload", "cat_notes", "cat_obs_fields",
               "cat_tags","cat_projects", "created_at_year", "observed_on_year",
               "created_at_month",
               "observed_on_month", "country","kingdom", "phylum", "class", "order", "family", "genus","taxon_rank",
               "sum_pixels")
               
```

There were issues with some observations that got NA in the sum of pixels of all photos because at least 1 observation was not available. I chose to remove these obs (at least for now) because if I fix it, it won't match n_photos that I calculated.

e.g: observation has nested dataset whith 4 rows of photos so n_photos was counted as 4. However, one of them is not available (unclear why) so only 3 have actual information, and pixels come out as NA. This happened in 21 observations, and I also removed one more observation that has no photos (even though that shouldn't be possible).

```{r}
data <- tidy_data %>% 
        filter(!is.na(sum_pixels) & n_photos != 0) %>% 
        mutate(mean_pixels = sum_pixels/n_photos,
               log_photos = log10(n_photos),
               log_sp = log10(sum_pixels),
               log_mp = log10(mean_pixels),
               rg=as.numeric(quality_grade == "research"))

```

## Model for number of photos

```{r}
percentage_data <- data %>% 
        mutate(log_photos=round(log_photos,1)) %>% 
        group_by(log_photos) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 
       

ggplot(percentage_data, aes(x = log_photos, y = percentage, size=total_count)) +
        geom_point() +
        theme_minimal()
```

```{r}
fit_n_photos<- glm(rg ~ log_photos, data=data, family = binomial)

summary(fit_n_photos)
```

```{r}
sjPlot::plot_model(fit_n_photos,type="pred", terms = "log_photos") +
        scale_y_continuous(limits=c(0,1))+
        theme_minimal()
```

## Model for pixels

### Sum of pixels of all photos in observation

```{r}
ggplot(data,aes(x = sum_pixels))+
        geom_histogram(bins=100)+
        scale_x_continuous(labels = scales::label_number())+
        labs(title = "Histogram sum of pixels bins=100")+
        theme_minimal()
```

I also tried log10 of total sum of pixels per observation:

```{r}
ggplot(data,aes(x = log_sp))+
        geom_histogram(bins=100)+
        scale_x_continuous(labels = scales::label_number())+
        labs(title = "Histogram log10 of sum of pixels bins=100")+
        theme_minimal()
```

```{r}

percentage_data <- data %>% 
        group_by(log_sp) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 
       

ggplot(percentage_data, aes(x = log_sp, y = percentage, size=total_count)) +
        geom_point() 
```

I tried mutate(log_sp=round(log_sp,1)) because we've done before - but help! I don't really understand it :)

```{r}

percentage_data <- data %>% 
        mutate(log_sp=round(log_sp,1)) %>% 
        group_by(log_sp) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 
       

ggplot(percentage_data, aes(x = log_sp, y = percentage, size=total_count)) +
        geom_point() 

```

```{r}
fit_sum_pixels<- glm(rg ~ log_sp, data=data,family = binomial)
#look at intercept and coefficients
summary(fit_sum_pixels)
```

```{r}
sjPlot::plot_model(fit_sum_pixels, type="pred", terms = "log_sp") +
        scale_y_continuous(limits=c(0,1))+
        theme_minimal()
```

### Mean of pixels per observation

```{r}

ggplot(data,aes(x = mean_pixels))+
        geom_histogram(bins=100)+
        scale_x_continuous(labels = scales::label_number())+
        labs(title = "Histogram mean of pixels bins=100")+
        theme_minimal()
```

```{r}

percentage_data <- data %>% 
        mutate(log_mp=round(log_mp,1)) %>% 
        group_by(log_mp) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 
       

ggplot(percentage_data, aes(x = log_mp, y = percentage, size=total_count)) +
        geom_point() 


```

```{r}
fit_mean_pixels<- glm(rg ~ log_mp, data=data,family = binomial)

#look at intercept and coefficients
summary(fit_mean_pixels)
```

```{r}
sjPlot::plot_model(fit_mean_pixels, type="pred", terms = "log_mp") +
        scale_y_continuous(limits=c(0,1))+
        theme_minimal()
```
