---
title: "EDA samples"
format:
  html:
    toc: true           # Enables the table of contents
    toc-location: body
editor: visual
---

After overall model...

To investigate the apparent counter intuitive relationship observed for the number of photos, we took a random sample of a total of 200 observations across four groups: 50 observations each for combinations of Quality Grade (RG or Needs ID) and number of photos (1 or 6). We then manually inspected these observations to identify possible reasons for differences in being identified to Research Grade.

It would be expected that 6 photos would be better than 1. However, in following plot we can see that there is a downward tendency between 1 and 6. For more than 6 photos the percentage started increasing, as expected. We therefore selected the tail ends of the relationship we were not expecting, 1 and 6.

```{r}
#| code-fold: true
#| output: false
library(tidyverse)
data <- read_csv("../data/annotated_samples.csv")

data <- data %>% 
        unite("quality_n_photos", quality_grade, n_photos, sep = "_", remove = FALSE) 

data$evaluation <- factor(data$evaluation, levels = c("bad", "average", "good", "great"))

```

```{r}
#| code-fold: true
#| output: false
tidy_data <- read_csv("../data/tidy_data/tidy_data_verifiable.csv")

```

```{r}
#| code-fold: true

percentage_data <- tidy_data %>% 
        select(n_photos, quality_grade) %>% 
        filter(n_photos != 0) %>% 
        mutate(log_photos = log10(n_photos),
               rg=as.numeric(quality_grade == "research")) %>% 
        mutate(log_photos=round(log_photos,1)) %>% 
        group_by(log_photos) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 

rm(tidy_data)

ggplot(percentage_data, aes(x = log_photos, y = percentage, size=total_count)) +
        geom_point() +
        labs(x = "log10 number of photos", 
             y = "RG %")+
        theme_minimal()
```

Aspects I classified:

-   growth_form - write how i got the categories in methods

-   evaluation

    -   bad - difficult to differentiate important characteristics (e.g. photo of top of cap only, blurry photos, can see side but not underside)
    -   average - "normal" observation (e.g. photo from the side but can see some detail, no close ups, a little blurry, can't check substrate)
    -   good - can clearly see different parts of fungi and details, either through close ups or there is good enough resolution to zoom-in. May include close ups of hymenium, photos out of soil, cut or presence of latex
    -   great - characteristics of good + microscopy or chemical tests

-   majority_similar - majority of photos in the observation were similar (y/n/not applicable (for 1 photo))

-   microscopy_photos - observation has microscopy photos (y/n)

-   chemical_test - photos or notes on chemical tests (not including microscopy reagents) (y/n)

    -   note: no observation in any group had chemical tests

-   details - zoom in on details, can see details of the underside of mushroom, details on notes (y/n)

-   cut_latex_ink_sporeprint - (y/n)

-   substrate_visible - (y/n)

-   interacted_with - has received comment or ID from another user

-   agreed_higher_level - community has agreed on a higher level, e.g. at least genus is right OR there was disagreement but other people have agreed

-   observer_ID_matches - 1st ID was made by observer matches community ID at species or genus level (y_species/y_genus/NA )

-   n_joined_projects - includes non traditional projects (number that shows up on website)

-   old_bad_quality_specimen

## Evaluation of the quality of the observation

In general, observations have a better quality when they have more photos (higher percentage of above average observations). Higher number allows user to capture different angles and details, even if inadvertently and can sometimes compensate for poor quality photos. No observations in 1 photo samples had microscopy or chemical tests so none were classified as great.

Research 1 has a lot of bad observations - these are either wrongly applied RGs or taxa that is very easily identifiable, even with very little detail.

```{r}
#| code-fold: true

count_eval<- data %>%
        group_by(quality_grade, n_photos) %>% 
        count(evaluation)

ggplot(count_eval, aes(x = evaluation, y = n, fill = evaluation)) +
        geom_bar(stat = "identity") +
        geom_text(aes(label = n), vjust = -0.5)+
        facet_wrap(quality_grade ~ n_photos)+
        theme_minimal()
```

```{r}
#| code-fold: true

data %>%
  ggplot(aes(x = quality_n_photos, fill = evaluation)) +
        geom_bar(position = "fill") +
        ylab("Proportion of observations")+
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme_minimal()
```

## Growth Forms

Number of photos needed to reach RG seems to be related to the morphology of fungi.

As expected, pileate-stipitate fungi are well represented across all sample groups.

**All groups:**

```{r}
#| code-fold: true
ggplot(data, aes(x = growth_form, fill = growth_form)) +
        geom_bar() +
        geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5)+
        theme_minimal()+
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**By sample group:**

Cap and stem mushrooms were more represented in needs ID samples.

The percentage of bracket fungi observations is higher for RG groups, especially with only one photo.

More Xylarioid observations in Research 1 as well.

Not much difference for lichens.

This might be influenced by over representation of certain species in specific samples. I added count of scientific name in observation below.

E.g. in RG1 there are 5 counts of Xylaria hypoxylon, 3 of Fomes fomentarius and other bracket fungi, also Amanita. This doesn't happen for RG6, where there are no duplicates of scientific names.

So the higher proportion of Xylarioids in RG1 is not necessarily because that is an easier growth form to document, but because Xylaria hypoxylon is overepresented.

The counts for needs id 1 photo has more counts of higher ranks like Agaricales and Fungi, more so than needs ID 6 that has duplicates at genus level.

```{r}
#| code-fold: true

count_gf<- data %>%
        group_by(quality_grade, n_photos) %>% 
        count(growth_form)

ggplot(count_gf, aes(x = growth_form, y = n, fill = growth_form)) +
        geom_col() +
        geom_text(aes(label = n), vjust = -0.5)+
        facet_wrap(quality_grade ~ n_photos)+
        theme_minimal()+
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| code-fold: true
# 
# data %>%
#   ggplot(aes(x = quality_n_photos, fill = growth_form)) +
#         geom_bar(position = "fill") +
#         theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#         theme_minimal()

```

```{r}
#| code-fold: true
data_percent <- data %>%
          group_by(quality_n_photos, growth_form) %>%
          summarise(n = n(), .groups = "drop") %>%
          group_by(quality_n_photos) %>%
          mutate(percentage = n / sum(n))

# Plot
ggplot(data_percent, aes(x = quality_n_photos, y = percentage, fill = growth_form)) +
          geom_col(position = "dodge") +
          scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
          theme_minimal() +
          theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

From plots, there seemed to be less variety of forms for RG1, compared to the others, namely RG 6. Those might be the growth forms for which it is very difficult to get all needed info in just one photo. But this might not be significant:

```{r}
#| code-fold: true

data %>% 
        group_by(quality_n_photos) %>%
        summarise(n_growth_forms = n_distinct(growth_form), .groups = "drop") %>% 
        arrange(desc(n_growth_forms)) %>% 
        knitr::kable()
```

## Taxonomy

#### Taxonomic rank: Phylum

Slightly higher proportion for Ascomycota in RG1.

```{r}
data %>%
  ggplot(aes(x = quality_n_photos, fill = fct_infreq(phylum))) +
        geom_bar(position = "fill") +
        ylab("Proportion of observations")+
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme_minimal()
```

```{r}
#| code-fold: true
# count_phylum<- data %>%
#         group_by(quality_grade, n_photos) %>% 
#         count(phylum)
# 
# ggplot(count_phylum, aes(x = fct_infreq(phylum), y = n, fill = phylum)) +
#         geom_bar(stat = "identity") +
#         facet_wrap(quality_grade ~ n_photos)+
#         theme_minimal()+
#         theme(axis.text.x = element_text(angle = 45, hjust = 1),
#               legend.position = "none")
```

#### Taxonomic rank: Order

Needs id 1 had 5 observations that had not identified to order level. This didn't happen with needs id 6 which suggests that adding photos is at least beneficial to reaching a lower rank id.

Agaricales is more represented in NI6 - matches growth forms. Might just be because there are a lot of confounding species - might be difficult even with more photos.

More polyporales and Xylariales in RG1 - matches growth forms.

More Russulales at RG6.

```{r}
#| code-fold: true
# 
# data %>%
#   ggplot(aes(x = quality_n_photos, fill = fct_infreq(order))) +
#         geom_bar(position = "fill") +
#         ylab("Proportion of observations")+
#         theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#         theme_minimal()


```

```{r}
#| code-fold: true

count_order<- data %>%
        group_by(quality_grade, n_photos) %>% 
        count(order)

ggplot(count_order, aes(x = fct_infreq(order), y = n, fill = order)) +
        geom_bar(stat = "identity") +
        facet_wrap(quality_grade ~ n_photos)+
        theme_minimal()+
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              legend.position = "none")
```

```{r}
#| code-fold: true
data_percent <- data %>%
          group_by(quality_n_photos, order) %>%
          summarise(n = n(), .groups = "drop") %>%
          group_by(quality_n_photos) %>%
          mutate(percentage = n / sum(n))

# Plot
ggplot(data_percent, aes(x = quality_n_photos, y = percentage, fill = order)) +
        geom_col(position = "dodge") +
        scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
        theme_minimal()
        
```

#### Taxonomic rank: Family

There were 3 obs not identified to family rank for Needs ID 6, and 13 for Needs ID 1, which again suggests that adding photos is at least beneficial to reaching a lower rank id.

RG1: Amanitaceae, Xylariaceae and Polyporaceae overepresented, which matches Growth forms.

NI 1: Not identified dominates (26%)

NI 6: More e.g. Agaricaceae

```{r}
#| code-fold: true
#| fig.height: 10
#| fig.width: 17
count_family<- data %>%
        group_by(quality_grade, n_photos) %>% 
        count(family)

ggplot(count_family, aes(x = fct_infreq(family), y = n, fill = family)) +
        geom_bar(stat = "identity") +
        facet_wrap(quality_grade ~ n_photos)+
        geom_text(aes(label = n), vjust = -0.5)+
        theme_minimal()+
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              legend.position = "none")
```

There was more variety of families for groups that had 6 photos.

```{r}
#| code-fold: true

data %>% 
        group_by(quality_n_photos) %>%
        summarise(n_families = n_distinct(family), .groups = "drop") %>% 
        arrange(desc(n_families)) %>% 
        knitr::kable()
```

#### Count of scientific names - are there prevalent species in certain n_photos/ quality grade combinations?

```{r}
#| code-fold: true

 data %>%
         group_by(quality_n_photos, scientificName) %>%
         summarise(count = n(), .groups = "drop") %>% 
         arrange(quality_n_photos, desc(count)) %>% 
         knitr::kable()
```

## Other details

### Majority is similar

Whether the majority of photos in the observation were similar between them, meaning that even if you had 6 photos, they were mostly redundant.

Results seem counter intuitive to me. There was more redundancy in RG observations with 6 photos than needs ID with 6 photos. Might just be because the species were inherently easier to identify, so having a lot of photos was not important?

```{r}
#| code-fold: true

data %>%
  ggplot(aes(x = quality_n_photos, fill = majority_similar)) +
        geom_bar(position = "fill") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme_minimal()
```

### Microscopy photos

To have microscopy photos you generally have more than 1. But there isn't a difference between needs_id 6 and research 6.

```{r}
#| code-fold: true

data %>%
  ggplot(aes(x = quality_n_photos, fill = microscopy_photos)) +
        geom_bar(position = "fill") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme_minimal()
```

### Details

What it means: zoom in on details, can see details of the underside of mushroom, details on notes

More photos means that you often capture more details, even inadvertently.

As expected, for the same number of photos, % of observations that had details was higher for Research Grade samples. This suggests that capturing details likely increases your chances of reaching RG? (Not sure I can say this as conclusion)

However - a lot of observations with 6 photos that are needs id have details -suggesting that it's not necessarily a problem of capturing detail, but rather what is being observed.

```{r}
#| code-fold: true
data %>%
  ggplot(aes(x = quality_n_photos, fill = details)) +
        geom_bar() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme_minimal()
```

### cut_latex_ink_sporeprint

To capture these things, you often need more photos so this is more prevalent for samples with 6 photos. For the same number of photos % that these details was the same for needs id and research samples,

Therefore it doesn't seem make a difference in reaching RG, even if we know that for some species these are important morphological characteristics.

```{r}
#| code-fold: true
data %>%
  ggplot(aes(x = quality_n_photos, fill = cut_latex_ink_sporeprint)) +
        geom_bar() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme_minimal()
```

### Substrate visible

Typically visible, RG observations with 6 photos have the smallest proportion of observations where it is not visible.

```{r}
#| code-fold: true
data %>%
  ggplot(aes(x = quality_n_photos, fill = substrate_visible)) +
        geom_bar() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme_minimal()
```

### Interacted with

Sample of needs id observations with 6 photos had low proportion of observations that had been interacted with (either comments or id suggestions). Either they are difficult taxa so people tend to not engage or they are among the "forgotten species". Discrepancy between 1 and 6 photos isn't clear to me.

```{r}
#| code-fold: true
data %>%
  ggplot(aes(x = quality_n_photos, fill = interacted_with)) +
        geom_bar() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme_minimal()
```

### Agreed at higher level

Not much difference.

```{r}
#| code-fold: true
data %>%
  ggplot(aes(x = quality_n_photos, fill = agreed_higher_level)) +
        geom_bar() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme_minimal()
```

### Observer ID matches community ID

For needs id:

-   not applicable: there were no ids, hence no community id

-   research 6 has lower % of correct initial observer suggestion than needs id 6 - might be indicative that the taxa itself wasn't the most obvious, and computer vision wasn't correct or as helpful?

```{r}
#| code-fold: true
data %>%
  ggplot(aes(x = quality_n_photos, fill = observer_ID_matches)) +
        geom_bar() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme_minimal()
```

### Number of joined projects

Didn't conclude much - observations with 6 photos generally had more projects. Tail for RG samples had more projects.

```{r}
#| code-fold: true

count_joined_projects<- data %>%
         count(quality_n_photos, n_joined_projects)

ggplot(count_joined_projects, aes(x = fct_infreq(as.factor(n_joined_projects)), y = n)) +
        geom_col() +
        geom_text(aes(label = n), vjust = -0.5)+
        facet_wrap(~ quality_n_photos)+
        theme_minimal()+
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              legend.position = "none")
```

### Old or bad quality specimens

Needs ID 6 had higher number of observations with bad quality specimens, still not very high though.

```{r}
#| code-fold: true
data %>%
  ggplot(aes(x = quality_n_photos, fill = old_bad_quality_specimen)) +
        geom_bar() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme_minimal()
```
