---
title: "EDA observation characteristics - notes, observation fields, projects and tags"
format: html
editor: visual
---

```{r}
#| label: load packages and data
#| code-fold: true
#| output: false
library(tidyverse)


tidy_data <- read.csv("../data/tidy_data/tidy_data_verifiable.csv")
rank_order <- readRDS("rank_order.RDS")
tidy_data <- tidy_data %>%
        mutate(taxon_rank = factor(taxon_rank, levels = rank_order, ordered = TRUE))

tidy_data <- tidy_data %>% 
        filter(!is.na(sum_pixels) & n_photos != 0) %>% 
        mutate(mean_pixels = sum_pixels/n_photos,
               log_photos = log10(n_photos),
               log_sp = log10(sum_pixels),
               log_mp = log10(mean_pixels),
               rg=as.numeric(quality_grade == "research"),
               log_dsu = log10(days_since_upload + 1),
               log_notes = log10(length_notes + .Machine$double.eps),
               log_obs_fields = log10(n_obs_fields + .Machine$double.eps),
               log_tags = log10(n_tags + .Machine$double.eps),
               log_projects = log10(n_projects + .Machine$double.eps),
               )

```

## Notes

#### 1. Presence vs absence 

```{r}
#| label: fig presence notes
#| code-fold: true
ggplot(tidy_data, aes(x = presence_notes)) +
        geom_bar() +
        labs(x = "Presence of notes", 
             y = "Number of observations")+
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal() 

ggplot(tidy_data, aes(x = presence_notes, fill=quality_grade)) +
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Presence of notes", 
             y = "Number of observations")+
        theme_minimal()

```

#### 2. Note length - number of characters

```{r}
percentage_data <- tidy_data %>%
        group_by(length_notes) %>% 
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 

ggplot(percentage_data, aes(x = length_notes, y = percentage, size=total_count)) +
        geom_point() +
        theme_minimal()
```

Log transformation:

Trying out log10+1 just to visualise data. Tried with and without rounding:

```{r}
percentage_data <- tidy_data %>%
        mutate(log_notes = log10(length_notes + 1)) %>% 
        group_by(log_notes) %>% 
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 

ggplot(percentage_data, aes(x = log_notes, y = percentage, size=total_count)) +
        geom_point() +
        theme_minimal()

```

```{r}
percentage_data <- tidy_data %>%
        mutate(log_notes = log10(length_notes),
               log_notes = round(log_notes,1)) %>% 
        group_by(log_notes) %>% 
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 

ggplot(percentage_data, aes(x = log_notes, y = percentage, size=total_count)) +
        geom_point() +
        theme_minimal()
```

changing log transformation to log10 + machine epsilon (transformation I have in the model)

```{r}
percentage_data <- tidy_data %>%
        mutate(log_notes_eps = log10(length_notes + .Machine$double.eps),
               log_notes_eps = round(log_notes_eps,1)) %>% 
        group_by(log_notes_eps) %>% 
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 

ggplot(percentage_data, aes(x = log_notes_eps, y = percentage, size=total_count)) +
        geom_point() +
        theme_minimal()
```

## Observation Fields

#### 1. Presence vs absence

```{r}
#| label: figs presence obs fields
#| code-fold: true
#| fig-height: 6

tidy_data <- tidy_data %>%
        mutate(presence_obs_fields = ifelse(n_obs_fields == 0, "no", "yes"))

ggplot(tidy_data, aes(x = presence_obs_fields)) +
        geom_bar() +
        labs(x = "Presence of observation fields", 
             y = "Number of observations")+
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal()
        
```

```{r}
#| label: figs presence obs fields ~ quality grade
#| code-fold: true

ggplot(tidy_data, aes(x = presence_obs_fields, fill=quality_grade)) +
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Presence of observation fields", 
             y = "Number of observations")+
        theme_minimal()
        
```

#### 2. Number of observation fields

```{r}
percentage_data <- tidy_data %>%
        group_by(n_obs_fields) %>% 
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 

ggplot(percentage_data, aes(x = n_obs_fields, y = percentage, size=total_count)) +
        geom_point() +
        theme_minimal()
```

Trying with log10 + 1. However, data is not that scattered - is it really needed?

```{r}
percentage_data <- tidy_data %>%
        mutate(log_obs_fields = log10(n_obs_fields + 1),
               log_obs_fields = round(log_obs_fields,1)) %>% 
        group_by(log_obs_fields) %>% 
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 

ggplot(percentage_data, aes(x = log_obs_fields, y = percentage, size=total_count)) +
        geom_point() +
        theme_minimal()
```

Makes sense because most observations (even excluding 0) have few observation fields: (?)

```{r}
ggplot(tidy_data,aes(x = n_obs_fields))+
        geom_bar()+
        labs(x = "Number of observation fields", 
             y = "Number of observations (log10)")+
        scale_y_log10()+
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal()
```

## Projects

#### 1. Presence vs absence

```{r}
#| label: figs presence in projects
#| code-fold: true
tidy_data <- tidy_data %>%
        mutate(presence_projects = ifelse(n_projects == 0, "no", "yes"))

ggplot(tidy_data, aes(x = presence_projects)) +
        geom_bar()+
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal() 
```

```{r}
#| label: figs presence in projects ~ quality grade
#| code-fold: true
ggplot(tidy_data,aes(x = presence_projects, fill= quality_grade))+
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        theme_minimal()  
```

#### 2. Number of projects

```{r}
#| code-fold: true
ggplot(tidy_data,aes(x = n_projects))+
        geom_bar()+
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal()
```

```{r}
percentage_data <- tidy_data %>%
        group_by(n_projects) %>% 
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 

ggplot(percentage_data, aes(x = n_projects, y = percentage, size=total_count)) +
        geom_point() +
        theme_minimal()
```

log10+1 with rounding:

```{r}
percentage_data <- tidy_data %>%
        mutate(log_n_projects = log10(n_projects + 1),
               log_n_projects = round(log_n_projects,1)) %>% 
        group_by(log_n_projects) %>% 
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 

ggplot(percentage_data, aes(x = log_n_projects, y = percentage, size=total_count)) +
        geom_point() +
        theme_minimal()
```

### Microscopy

```{r}
#| label: fig microscopy_project
#| code-fold: true
ggplot(tidy_data, aes(x = microscopy_project)) +
        geom_bar() +
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal() 
```

```{r}
#| label: fig microscopy_project ~ quality grade
#| code-fold: true
ggplot(tidy_data, aes(x = microscopy_project, fill = quality_grade)) +
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        theme_minimal()
```

## Tags

#### 1. Presence vs absence

```{r}
#| label: fig presence tags
#| code-fold: true
tidy_data <- tidy_data %>%
        mutate(presence_tags = ifelse(n_projects == 0, "no", "yes"))

ggplot(tidy_data, aes(x = presence_tags)) +
        geom_bar() +
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal() 

ggplot(tidy_data,aes(x = presence_tags, fill= quality_grade))+
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Number of projects", 
             y = "Proportion", 
             title = "Number of projects ~ Quality grade")+
        theme_minimal() 
```

#### 2. Number of tags

```{r}
#| label: fig number tags
#| code-fold: true

ggplot(tidy_data,aes(x = n_tags))+
        geom_bar()+
        labs(x = "Number of tags", 
             y = "Number of observations (log10)",
             title = "Number of tags")+
        scale_y_log10()+
        theme_minimal()


```

```{r}
percentage_data <- tidy_data %>%
        group_by(n_tags) %>% 
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 

ggplot(percentage_data, aes(x = n_tags, y = percentage, size=total_count)) +
        geom_point() +
        theme_minimal()
```

Trying with log10 + 1, with rounding:

```{r}
percentage_data <- tidy_data %>%
        mutate(log_n_tags = log10(n_tags + 1),
               log_n_tags = round(log_n_tags,1)) %>% 
        group_by(log_n_tags) %>% 
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 

ggplot(percentage_data, aes(x = log_n_tags, y = percentage, size=total_count)) +
        geom_point() +
        theme_minimal()
```
