---
title: "Exploratory Analysis - Time plots"
date: November 12 2024
format: html
editor: visual
---

iNat gives two types of dates+time of each observation. observed_on corresponds to the actual date of the observation (date when photo was taken) and created_at the date when the observation was uploaded. iNaturalist was created in 2008 and the first Fungi observation was uploaded in May 2009, but older photos have also since been uploaded so some observed_on values date further back.

```{r}
#| code-fold: true
#| output: false
library(tidyverse)
tidy_data <- read_csv("../data/tidy_data/tidy_data_verifiable.csv")
```

```{r}
#| label: fig observations and uploads per year
#| code-fold: true

tidy_data %>%
        pivot_longer(cols = c(observed_on_year, created_at_year),
                     names_to = "year_type",
                     values_to = "year") %>% 
        ggplot(aes(x = year, fill = year_type)) +
        geom_bar(position = "dodge") +
        labs(title = "Observations and uploads per year",
             x = "Year", 
             y = "Count",
             fill = "Type") +
        theme_minimal() +
        scale_y_continuous(labels = scales::label_number())
```

```{r}
#| label: fig observations and uploads per year since 2008
#| code-fold: true

tidy_data %>%
        filter(observed_on >= 2008  ) %>% 
        pivot_longer(cols = c(observed_on_year, created_at_year),
                     names_to = "year_type",
                     values_to = "year") %>% 
        ggplot(aes(x = year, fill = year_type)) +
        geom_bar(position = "dodge") +
        labs(title = "Observations and uploads per year since 2008",
             x = "Year", 
             y = "Count",
             fill = "Type") +
        theme_minimal() +
        scale_y_continuous(labels = scales::label_number())
```

There seems to be a seasonal pattern to both observations and uploads:

```{r}
#| label: fig observations and uploads per month
#| code-fold: true

tidy_data <- tidy_data %>%
        mutate(observed_on_month = factor(observed_on_month, levels = month.abb, ordered = TRUE)) %>%
        mutate(created_at_month = factor(created_at_month, levels = month.abb, ordered = TRUE))

tidy_data %>%
        pivot_longer(cols = c(observed_on_month, created_at_month),
                     names_to = "month_type",
                     values_to = "month") %>% 
        ggplot(aes(x = month, fill = month_type)) +
        geom_bar(position = "dodge") +
        labs(title = "Observations and uploads per month",
             x = "Month", 
             y = "Count",
             fill = "Type") +
        theme_minimal() +
        scale_y_continuous(labels = scales::label_number(),
                           breaks = seq(50000, 600000, by = 100000))

```

```{r}
#| label: fig perc. of RG per day
#| code-fold: true

percentage_data <- tidy_data %>% 
        group_by(days_since_upload) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100)

# ggplot(percentage_data,aes(x = days_since_upload, y= percentage))+
#         geom_point(size = 0.5, alpha = 0.5) +
#         geom_line(linewidth = 0.2)+
#         labs(x = "Days since upload", y = "RG%", title = "Percentage of RG per day") +
#         theme_minimal()

ggplot(percentage_data,aes(x = days_since_upload, y= percentage))+
        geom_point(size = 0.5, alpha = 0.5) +
        geom_smooth()+
        labs(x = "Days since upload", y = "RG%", title = "Percentage of RG per day") +
        scale_x_continuous(breaks = seq(0, max(tidy_data$days_since_upload), by = 365))+
        theme_minimal()

```

Note: Contrarily to other plots the most recent observations are on the left of the plot.

Proportion of RG increased with the longer the observation stays on the platform. But there also seems to be a seasonal pattern to the % of research grade, which is more surprising. Possible explanations:

-   "off season" species are harder to identify or less known to users in the platform

-   there are less identifiers in the platform in the "off season" so you are less likely to get an ID

To see this, we differentiated between proportion of RG according to time of observation and upload separately:

```{r}
#| label: fig observations and uploads per month - quality grade
#| code-fold: true

tidy_data %>%
        pivot_longer(cols = c(observed_on_month, created_at_month),
                     names_to = "month_type",
                     values_to = "month") %>% 
        ggplot(aes(x = month, fill= quality_grade)) +
        geom_bar(position = "fill") +
        facet_wrap(~ month_type)+
        labs(title = "Observations and uploads per month ~ Quality grade",
             x = "Month", 
             y = "Proportion of RG",
             fill = "Type") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        scale_y_continuous(labels = scales::label_number())

```

Even though there seems to be a pattern with observations made or uploaded between June and November having less % RG, this doesn't seem to match up very well with the pattern in the previous plot it was more pronounced. To test whether this is because the increase in RG % has increased with time and it is overshadowing differences when they are all put together, I separated between different years:

```{r}
#| label: fig observations per month - quality grade
#| code-fold: true

tidy_data %>%
        ggplot(aes(x = observed_on_month, fill= quality_grade)) +
        geom_bar(position = "fill") +
        facet_wrap(~created_at_year)+
        labs(title = "Observations per month ~ Quality grade per year uploaded at",
             x = "Month", 
             y = "Proportion of RG",
             fill = "Type") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        scale_y_continuous(labels = scales::label_number())


tidy_data %>%
        ggplot(aes(x = observed_on_month, fill= quality_grade)) +
        geom_bar(position = "fill") +
        facet_wrap(~observed_on_year)+
        labs(title = "Observations per month ~ Quality grade, per year observed on",
             x = "Month", 
             y = "Proportion of RG",
             fill = "Type") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        scale_y_continuous(labels = scales::label_number())
```

```{r}
#| label: fig uploads per month - quality grade
#| code-fold: true
tidy_data %>%
        ggplot(aes(x = created_at_month, fill= quality_grade)) +
        geom_bar(position = "fill") +
        facet_wrap(~created_at_year)+
        labs(title = "Uploads per month ~ Quality grade, per year of upload",
             x = "Month", 
             y = "Proportion of RG",
             fill = "Type") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        scale_y_continuous(labels = scales::label_number())
```

```{r}
#| label: fig uploads per day
#| code-fold: true
tidy_data %>% 
        mutate(created_at = as.Date(created_at)) %>% 
        count(day= floor_date(created_at,"day")) %>% 
        ggplot(aes(x=day, y= n))+
        geom_line() +
        geom_smooth() +
        scale_x_date(breaks = "1 year",
                     date_minor_breaks ="6 months",
                     date_labels = "%Y") +
        labs(x = "t", y = "Count", title = " Uploads per day ")+
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: fig uploads per day after 2019
#| code-fold: true
tidy_data %>% 
        filter(created_at_year >= 2019) %>% 
        mutate(created_at = as.Date(created_at)) %>% 
        count(day= floor_date(created_at,"day")) %>% 
        ggplot(aes(x=day, y= n))+
        geom_line()+
        geom_smooth()+
        scale_x_date(breaks = "3 months",
                     date_minor_breaks ="1 month",
                     date_labels = "%b %Y") +
        labs(x = "t", y = "Count", title = " Uploads per day after 2019 ")+
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
