---
title: "Exploratory Analysis"
date: November 12 2024
format: html
editor: visual
editor_options: 
  chunk_output_type: inline
---

```{r}
#| label: load packages and data
#| code-fold: true
#| output: false
library(tidyverse)
library(viridis)
library(scales)
library(ggExtra)
library(patchwork)

tidy_data_ <- read.csv("../data/tidy_data/tidy_data_verifiable.csv")
rank_order <- readRDS("rank_order.RDS")
tidy_data <- tidy_data %>%
        mutate(taxon.rank = factor(taxon.rank, levels = rank_order, ordered = TRUE))
```

Large majority of observations belong to Basidiomycota and Ascomycota:

```{r}
#| label: fig phyla
#| code-fold: true
ggplot(tidy_data,aes(x = fct_infreq(phylum), fill= phylum))+
        geom_bar()+
        labs(title = "Number of observations per phylum",
             x = "Phylum", y = "Number of observations (log10)")+
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
        scale_y_log10(labels = scales::label_number())
```

```{r}
#| label: fig quality grade phylum
#| code-fold: true

data_summary_grade <- tidy_data%>%
        group_by(phylum, quality_grade) %>%
        summarise(count = n(), .groups = 'drop') #ungroups at the end

phylum_grade_order <- data_summary_grade %>%
        group_by(phylum) %>%
        summarise(prop_research = 
                          sum(count[quality_grade == 'research']) / sum(count)) %>% 
        arrange(desc(prop_research)) %>%
        pull(phylum)

# Reorder phylum and taxon rank as factors
tidy_data <- tidy_data %>%
        mutate(phylum = factor(phylum, levels = phylum_grade_order), 
               taxon.rank = factor(taxon.rank, levels = rank_order))

stacked_plot<- ggplot(tidy_data, aes(x = phylum, fill = quality_grade)) +
        geom_bar(position = "fill") +
        labs(title = "RG proportion per phylum",
             x = "Phylum", y = "Proportion of obs", fill = "Quality Grade") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        theme_minimal()+
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

hist_plot <- ggplot(tidy_data,aes(x = phylum))+
        geom_bar()+
        labs( x = NULL, y = "No.obs. (log10)")+
        theme_minimal() +
        theme(axis.text.x = element_blank())+
        scale_y_log10(labels = scales::label_number())
        
hist_plot / stacked_plot + plot_layout(heights = c(1, 2))
```

Taxonomic predictor - taxon observation belongs to has influence of identification success (intrinsic predictor). However, size of the taxon (e.g. number of species in the phylum) could be confounding factor.

NA values correspond to observations at Fungi level (with exception of fossilised EX species).

-   Though Ascomycota has less observations, proportion of RG is slightly higher.

-   The 3 phyla with higher %RG have less than 100 obs.

## Plots - Taxon Rank

Observations were grouped by the lowest taxon rank they were identified to. (aka the level they got stuck at :) )

```{r}
#| label: fig lowest taxon rank
#| code-fold: true

ggplot(tidy_data,aes(x = taxon.rank, fill= taxon.rank))+
        geom_bar()+
        labs(title = "Lowest taxonomic level",
             x = "Taxon rank", 
             y = "Number of observations") +
        scale_fill_viridis_d(option = "inferno") +
        theme_minimal()+
        theme(legend.position = "none") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Only observations that reach species rank or lower will automatically be upgraded to RG if the community reaches a consensus. However, it is possible to tag an observation as "as good as it can be" (e.g. microscopy is needed to confirm species, but it was not done) in which case it will become RG for the level the community agreement is at. This is not very frequent however:

```{r}
#| label: fig lowest taxon rank by grade
#| code-fold: true

ggplot(tidy_data,aes(x = taxon.rank, fill= taxon.rank))+
        geom_bar()+
        labs(title = "Lowest taxonomic level grouped by quality grade",
             x = "Taxon rank", 
             y = "Number of observations") +
        facet_wrap(~quality_grade,ncol=1,scales="free_y")+
        scale_fill_viridis_d(option="inferno") +
        theme_minimal()+
        theme(legend.position = "none") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              axis.title.y = element_text(margin = margin(r = 10)))+
        scale_y_continuous(labels = scales::label_number())

```

**To do:** check how many IDs observations have (IDs from user other than observer \>0) to see identify the potential *forgotten observations* - Have they not moved to RG due to disagreements or because no one has looked at them? Could vary between regions and taxonomy and help us identify hotspots of identification success.

```{r}
#| label: fig taxon rank proportion by phylum below above species level 
#| code-fold: true

tidy_data <- tidy_data %>% 
        mutate(species_level = case_when(
    taxon.rank >= "species" ~ "species level or below",
    taxon.rank < "species" ~ "above species level",
    TRUE ~ NA_character_ #TRUE isn't a consdition being tested but a catch all
  )) 

#order by proportion of RG 
data_summary_rank <- tidy_data %>% 
        group_by(phylum, species_level) %>%
        summarise(count = n(), .groups = 'drop')

phylum_species_order <- data_summary_rank %>%
        group_by(phylum) %>%
        summarise(prop_species = sum(count[species_level == "species level or below"]) / sum(count)) %>%
        arrange(desc(prop_species)) %>%
        pull(phylum)

tidy_data %>% 
        mutate(phylum = factor(phylum, levels = phylum_species_order)) %>% 
        ggplot(aes(x = phylum, fill = species_level)) +
                geom_bar(position = "fill") +
                labs(title = "Proportion of lowest taxon ranks per phylum (above/below species level)",
                     x = "Phylum", y = "Proportion of observations", fill = "Lowest taxon rank") +
                theme_minimal()+
                theme(axis.text.x = element_text(angle = 45, hjust = 1),
                      legend.margin = margin(t = 60 , l = 15))

```

```{r}
#| label: fig taxon rank proportion by phylum below above species level - RG
#| code-fold: true

tidy_data <- tidy_data %>% 
        mutate(species_level = case_when(
    taxon.rank >= "species" ~ "species level or below",
    taxon.rank < "species" ~ "above species level",
    TRUE ~ NA_character_ #TRUE isn't a consdition being tested but a catch all
  )) 

#order by proportion of RG 
data_summary_rank <- tidy_data %>% 
        filter(quality_grade == "research") %>% 
        group_by(phylum, species_level) %>%
        summarise(count = n(), .groups = 'drop')

phylum_species_order <- data_summary_rank %>%
        group_by(phylum) %>%
        summarise(prop_species = sum(count[species_level == "species level or below"]) / sum(count)) %>%
        arrange(desc(prop_species)) %>%
        pull(phylum)

tidy_data %>% 
        filter(quality_grade == "research") %>% 
        mutate(phylum = factor(phylum, levels = phylum_species_order)) %>% 
        ggplot(aes(x = phylum, fill = species_level)) +
                geom_bar(position = "fill") +
                labs(title = "Prop. of lowest taxon ranks per phylum (above/below species level) - only RG",
                     x = "Phylum", y = "Proportion of observations", fill = "Lowest taxon rank") +
                theme_minimal()+
                theme(axis.text.x = element_text(angle = 45, hjust = 1),
                      legend.margin = margin(t = 60 , l = 15))
```

For reference of number of observations:

```{r}
#| label: fig lowest taxon rank by phylum
#| code-fold: true
#| fig-height: 6

ggplot(tidy_data,aes(x = taxon.rank, fill= taxon.rank))+
        geom_bar()+
        labs(title = "Lowest taxonomic level by phylum",
             x = "Taxon rank", 
             y = "Number of observations") +
        facet_wrap(~phylum, ncol=2, scales="free_y")+
        scale_fill_viridis_d(option = "inferno") +
        theme_minimal()+
        theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))+
        scale_y_continuous(labels = scales::label_number())

tidy_data %>% filter(quality_grade == "research") %>% 
        ggplot(aes(x = taxon.rank, fill= taxon.rank))+
                geom_bar()+
                labs(title = "Lowest taxonomic level by phylum for Research Grade observations",
                     x = "Taxon rank", 
                     y = "Number of observations") +
                facet_wrap(~phylum, ncol=2, scales="free_y")+
                scale_fill_viridis_d(option = "inferno") +
                theme_minimal()+
                theme(legend.position = "none",
                axis.text.x = element_text(angle = 45, hjust = 1))+
                scale_y_continuous(labels = scales::label_number())

```

More detailed plots:

```{r}
#| label: fig taxon rank proportion by phylum
#| code-fold: true

#order by proportion of RG 
data_summary_rank <- tidy_data %>% 
        group_by(phylum, taxon.rank) %>%
        summarise(count = n(), .groups = 'drop')

phylum_rank_order <- data_summary_rank %>%
        group_by(phylum) %>%
        summarise(prop_species = sum(count[taxon.rank == 'species']) / sum(count)) %>%
        arrange(desc(prop_species)) %>%
        pull(phylum)

tidy_data %>% 
        mutate(phylum = factor(phylum, levels = phylum_rank_order)) %>% 
        mutate(taxon.rank = factor(taxon.rank, levels = rank_order)) %>% 
        ggplot(aes(x = phylum, fill = taxon.rank)) +
                geom_bar(position = "fill") +
                labs(title = "Proportion of lowest taxon ranks per phylum",
                     x = "Phylum", 
                     y = "Proportion of observations", 
                     fill = "Lowest taxon rank") +
                scale_fill_viridis_d(option = "turbo") +
                theme_minimal()+
                theme(axis.text.x = element_text(angle = 45, hjust = 1),
                      legend.margin = margin(t = 60 , l = 15))


```

```{r}
#| label: fig taxon rank proportion by phylum - RG
#| code-fold: true

#order by proportion of RG 
data_summary_rank <- tidy_data %>%
        filter(quality_grade == "research") %>% 
        group_by(phylum, taxon.rank) %>%
        summarise(count = n(), .groups = 'drop')

phylum_rank_order <- data_summary_rank %>%
        group_by(phylum) %>%
        summarise(prop_species = sum(count[taxon.rank == 'species']) / sum(count)) %>%
        arrange(desc(prop_species)) %>%
        pull(phylum)

tidy_data %>% 
        filter(quality_grade == "research") %>% 
        mutate(phylum = factor(phylum, levels = phylum_rank_order)) %>% 
        mutate(taxon.rank = factor(taxon.rank, levels = rank_order)) %>% 
        ggplot(aes(x = phylum, fill = taxon.rank)) +
        geom_bar(position = "fill") +
                labs(title = "Proportion of lowest taxon ranks per phylum - only RG",
                     x = "Phylum", y = "Proportion of observations",
                     fill = "Lowest taxon rank") +
                scale_fill_viridis_d(option = "turbo") +
                theme_minimal()+
                theme(axis.text.x = element_text(angle = 45, hjust = 1),
                      legend.margin = margin(t = 60 , l = 15))
```

## Plots - time

iNat gives two types of dates+time of each observation. observed_on corresponds to the actual date of the observation (date when photo was taken) and created_at the date when the observation was uploaded. iNaturalist was created in 2008 and the first Fungi observation was uploaded in May 2009, but older photos have also since been uploaded so some observed_on values date further back.

```{r}
#| label: fig observations and uploads per year
#| code-fold: true

tidy_data %>%
        pivot_longer(cols = c(observed_on_year, created_at_year),
                     names_to = "year_type",
                     values_to = "year") %>% 
        ggplot(aes(x = year, fill = year_type)) +
        geom_bar(position = "dodge") +
        labs(title = "Observations and uploads per year",
             x = "Year", 
             y = "Count",
             fill = "Type") +
        theme_minimal() +
        scale_y_continuous(labels = scales::label_number())
```

```{r}
#| label: fig observations and uploads per year since 2008
#| code-fold: true

tidy_data %>%
        filter(observed_on >= 2008  ) %>% 
        pivot_longer(cols = c(observed_on_year, created_at_year),
                     names_to = "year_type",
                     values_to = "year") %>% 
        ggplot(aes(x = year, fill = year_type)) +
        geom_bar(position = "dodge") +
        labs(title = "Observations and uploads per year since 2008",
             x = "Year", 
             y = "Count",
             fill = "Type") +
        theme_minimal() +
        scale_y_continuous(labels = scales::label_number())
```

There seems to be a seasonal pattern to both observations and uploads:

```{r}
#| label: fig observations and uploads per month
#| code-fold: true

tidy_data <- tidy_data %>%
        mutate(observed_on_month = factor(observed_on_month, levels = month.abb, ordered = TRUE)) %>%
        mutate(created_at_month = factor(created_at_month, levels = month.abb, ordered = TRUE))

tidy_data %>%
        pivot_longer(cols = c(observed_on_month, created_at_month),
                     names_to = "month_type",
                     values_to = "month") %>% 
        ggplot(aes(x = month, fill = month_type)) +
        geom_bar(position = "dodge") +
        labs(title = "Observations and uploads per month",
             x = "Month", 
             y = "Count",
             fill = "Type") +
        theme_minimal() +
        scale_y_continuous(labels = scales::label_number(),
                           breaks = seq(50000, 600000, by = 100000))

```

`{# {r} # #| label: fig prop. of quality grade per day # #| code-fold: true # ggplot(tidy_data,aes(x = days_since_upload, fill = quality_grade))+ #         geom_bar(position = "fill") + #         scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+ #         labs(x = "Days since upload", y = "Proportion", title = "Proportion of quality grade per day")+ #         theme_minimal()`

Note: Contrarily to other plots the most recent observations are on the left of the plot.

Proportion of RG increased with the longer the observation stays on the platform. But there also seems to be a seasonal pattern to the % of research grade, which is more surprising. Possible explanations:

-   "off season" species are harder to identify or less known to users in the platform

-   there are less identifiers in the platform in the "off season" so you are less likely to get an ID

To see this, we differentiated between proportion of RG according to time of observation and upload separately:

```{r}
#| label: fig observations and uploads per month - quality grade
#| code-fold: true

tidy_data %>%
        pivot_longer(cols = c(observed_on_month, created_at_month),
                     names_to = "month_type",
                     values_to = "month") %>% 
        ggplot(aes(x = month, fill= quality_grade)) +
        geom_bar(position = "fill") +
        facet_wrap(~ month_type)+
        labs(title = "Observations and uploads per month ~ Quality grade",
             x = "Month", 
             y = "Proportion of RG",
             fill = "Type") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        scale_y_continuous(labels = scales::label_number())

tidy_data %>%
        filter(created_at_year ==2022 | created_at_year ==2021) %>% 
        pivot_longer(cols = c(observed_on_month, created_at_month),
                     names_to = "month_type",
                     values_to = "month") %>% 
        ggplot(aes(x = month, fill= quality_grade)) +
        geom_bar(position = "fill") +
        facet_wrap(~ month_type)+
        labs(title = "Observations and uploads per month ~ Quality grade",
             x = "Month", 
             y = "Proportion of RG",
             fill = "Type") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        scale_y_continuous(labels = scales::label_number())


```

Even though there seems to be a pattern with observations made or uploaded between June and November having less % RG, this doesn't seem to match up very well with the pattern in the previous plot it was more pronounced. To test whether this is because the increase in RG % has increased with time and it is overshadowing differences when they are all put together, I separated between different years:

```{r}
#| label: fig observations per month - quality grade
#| code-fold: true

tidy_data %>%
        ggplot(aes(x = observed_on_month, fill= quality_grade)) +
        geom_bar(position = "fill") +
        facet_wrap(~created_at_year)+
        labs(title = "Observations per month ~ Quality grade per year uploaded at",
             x = "Month", 
             y = "Proportion of RG",
             fill = "Type") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        scale_y_continuous(labels = scales::label_number())


tidy_data %>%
        ggplot(aes(x = observed_on_month, fill= quality_grade)) +
        geom_bar(position = "fill") +
        facet_wrap(~observed_on_year)+
        labs(title = "Observations per month ~ Quality grade, per year observed on",
             x = "Month", 
             y = "Proportion of RG",
             fill = "Type") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        scale_y_continuous(labels = scales::label_number())
```

```{r}
#| label: fig uploads per month - quality grade
#| code-fold: true
tidy_data %>%
        ggplot(aes(x = created_at_month, fill= quality_grade)) +
        geom_bar(position = "fill") +
        facet_wrap(~created_at_year)+
        labs(title = "Uploads per month ~ Quality grade, per year of upload",
             x = "Month", 
             y = "Proportion of RG",
             fill = "Type") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))+
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        scale_y_continuous(labels = scales::label_number())
```

```{r}
#| label: fig uploads per day
#| code-fold: true
tidy_data %>% 
        mutate(created_at = as.Date(created_at)) %>% 
        count(day= floor_date(created_at,"day")) %>% 
        ggplot(aes(x=day, y= n))+
        geom_line() +
        geom_smooth() +
        scale_x_date(breaks = "1 year",
                     date_minor_breaks ="6 months",
                     date_labels = "%Y") +
        labs(x = "t", y = "Count", title = " Uploads per day ")+
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: fig uploads per day after 2019
#| code-fold: true
tidy_data %>% 
        filter(created_at >= 2019) %>% 
        mutate(created_at = as.Date(created_at)) %>% 
        count(day= floor_date(created_at,"day")) %>% 
        ggplot(aes(x=day, y= n))+
        geom_line()+
        geom_smooth()+
        scale_x_date(breaks = "3 months",
                     date_minor_breaks ="1 month",
                     date_labels = "%b %Y") +
        labs(x = "t", y = "Count", title = " Uploads per day after 2019 ")+
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: fig perc. of RG per day
#| code-fold: true

percentage_data <- tidy_data %>% 
        group_by(days_since_upload) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100)

# ggplot(percentage_data,aes(x = days_since_upload, y= percentage))+
#         geom_point(size = 0.5, alpha = 0.5) +
#         geom_line(linewidth = 0.2)+
#         labs(x = "Days since upload", y = "RG%", title = "Percentage of RG per day") +
#         theme_minimal()

ggplot(percentage_data,aes(x = days_since_upload, y= percentage))+
        geom_point(size = 0.5, alpha = 0.5) +
        geom_smooth()+
        labs(x = "Days since upload", y = "RG%", title = "Percentage of RG per day") +
        theme_minimal()

```

## Number of photos

```{r}
#| label: fig photos per observation
#| code-fold: true
ggplot(tidy_data,aes(x = n_photos))+
        geom_histogram(binwidth = 1)+
        labs(x = "Number of photos", 
             y = "Number of observations",
             title = "Number of photos per observation")+
        scale_x_continuous(breaks = seq(0, 75, by = 5))+
        scale_y_continuous(labels = scales::label_number())+
        theme_minimal()

ggplot(tidy_data,aes(x = n_photos))+
        geom_histogram(binwidth = 1)+
        labs(x = "Number of photos", 
             y = "Number of observations (log10)",
             title = "Number of photos per observation")+
        scale_x_continuous(breaks = seq(0, 75, by = 5))+
        scale_y_log10(labels = scales::label_number())+
        theme_minimal()

```

```{r}
#| label: fig n photos ~ quality grade
#| code-fold: true

ggplot(tidy_data,aes(x = n_photos, fill = quality_grade))+
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Number of photos", 
             y = "Proportion", 
             title = "Proportion of quality grade ~ number of photos")+
        theme_minimal() 
```

```{r}
#| label: fig n photos ~ RG percentage
#| code-fold: true

percentage_data <- tidy_data %>% 
        group_by(n_photos) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 
       

ggplot(percentage_data, aes(x = n_photos, y = percentage, size= total_count)) +
        geom_point() +
        geom_smooth() +
        labs(x = "Number of photos", 
             y = "RG %",
             title = "Number photos ~ % RG (method = loess)") +
        scale_x_continuous(trans="log10",breaks = seq(0, 75, by = 5)) +
        theme_minimal()

#made research binary (research = 1 or 0) and transformed to log10 for x
tidy_data<- tidy_data %>% 
        mutate(rg=as.numeric(quality_grade == "research"),
               log_photos = log10(n_photos))

#first model for log_photos and phylum (additive only) (family is binomial because there are two possible outcomes 1 or 0)
#we filter because we're getting error because of infinite values - there is one value where nphotos = 0 and log(0) is undefined 
fit_n_photos_1<- glm(rg ~ log_photos + phylum, data=tidy_data %>% 
                             filter(!is.infinite(log_photos)), family = binomial)
summary(fit_n_photos_1)

#visualise model outputs
sjPlot::plot_model(fit_n_photos_1,type="pred", terms = c("log_photos", "phylum")) +
        scale_y_continuous(limits=c(0,1))

#random sample for testing
fit_data <- tidy_data %>% 
        sample_n(size=100000)

#second model with interaction between photos and phylum
fit_n_photos_2<- glm(rg ~ log_photos + phylum + log_photos:phylum, data=fit_data %>% 
                             filter(!is.infinite(log_photos)), family = binomial)
summary(fit_n_photos_2)
sjPlot::plot_model(fit_n_photos_2,type="pred", terms = c("log_photos", "phylum")) +
        scale_y_continuous(limits=c(0,1))

#sense check visualisation - does the model match the data sensibly
percentage_data <- tidy_data %>% 
        mutate(log_photos=round(log_photos,1)) %>% 
        group_by(log_photos) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) 
       

ggplot(percentage_data, aes(x = log_photos, y = percentage, size=total_count)) +
        geom_point() +
        labs(x = "Number of photos", 
             y = "RG %",
             title = "Number photos ~ % RG (method = loess)") +
        theme_minimal()
#answer: did not match (logistic model had a downward slope, but data goes down then up)

#try smth non linear, with gam
fit_n_photos_3<- mgcv::gam(rg ~ s(log_photos) + phylum, data=fit_data %>% 
                             filter(!is.infinite(log_photos)), family = binomial)
summary(fit_n_photos_3)
sjPlot::plot_model(fit_n_photos_3,type="pred", terms = c("log_photos", "phylum")) +
        scale_y_continuous(limits=c(0,1))

#answer: model matches the data but not a lot of meaningful variation
```

```{r}
#| label: fig n photos ~ RG percentage plus histogram
#| code-fold: true
percentage_data <- tidy_data %>% 
        group_by(n_photos) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) %>% 
        filter(n_photos != 0 & n_photos <= 30)

line_plot <- ggplot(percentage_data, aes(x = n_photos, y = percentage)) +
        geom_point() +
        geom_smooth() +
        labs(x = "Number of photos", 
             y = "RG %",
             title = "Number photos ~ % RG (method = loess)") +
        theme_minimal()

hist_plot <- tidy_data %>% 
        filter(n_photos != 0 & n_photos <= 30) %>% 
        ggplot(aes(x = n_photos))+
        geom_histogram(binwidth=1)+
        labs(x = "Number of photos", 
             y = "Number of observations",
             title = "Number of photos per observation")+
        scale_y_continuous(labels = scales::label_number())+
        geom_text(stat = "count", aes(label = after_stat(count)), 
                  vjust = -1,
                  size= 2)+
        theme_minimal()

percentage_data <- tidy_data %>% 
        group_by(n_photos) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) %>% 
        filter(n_photos != 0 & n_photos <= 20)

line_plot_2 <- ggplot(percentage_data, aes(x = n_photos, y = percentage)) +
        geom_point() +
        geom_smooth() +
        labs(x = "Number of photos", 
             y = "RG %",
             title = "Number photos ~ % RG (method = loess)") +
        theme_minimal()
hist_plot / line_plot / line_plot_2
```

```{r}
#| label: fig n photos ~ RG percentage ~ microscopy_project
#| code-fold: true

percentage_data <- tidy_data %>% 
        group_by(n_photos, microscopy_project) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100) %>% 
        filter(n_photos != 0 & n_photos <= 30) 

ggplot(percentage_data, aes(x = n_photos, y = percentage, colour= microscopy_project )) +
        geom_point(size= 0.5) +
        geom_smooth(linewidth= 0.5) +
        labs(x = "Number of photos", 
             y = "RG %",
             title = "Number photos ~ % RG (method = loess)") +
        scale_x_continuous(breaks = seq(0, 75, by = 10)) +
        theme_minimal()

ggplot(percentage_data, aes(x = n_photos, y = percentage, colour= microscopy_project )) +
        geom_point(size= 0.5) +
        geom_smooth(linewidth= 0.5) +
        facet_wrap(~microscopy_project)+
        labs(x = "Number of photos", 
             y = "RG %",
             title = "Number photos ~ % RG (method = loess)") +
        scale_x_continuous(breaks = seq(0, 75, by = 10)) +
        theme_minimal()
```

## Notes

```{r}
#| label: fig presence notes
#| code-fold: true
ggplot(tidy_data, aes(x = presence_notes)) +
        geom_bar() +
        labs(x = "Presence of notes", 
             y = "Number of observations", 
             title = " Presence of notes ")+
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal() 

ggplot(tidy_data, aes(x = presence_notes, fill=quality_grade)) +
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Presence of notes", 
             y = "Number of observations", 
             title = " Presence of notes ~ quality grade ")+
        theme_minimal()

```

```{r}
#| label: figs length notes
#| code-fold: true

tidy_data %>% 
        ggplot(aes(x = length_notes)) +
        labs(x = "Note length (number of characters)", 
             y = "Number of observations (log10)", 
             title = "Length of notes")+
        geom_bar() +
        scale_y_log10()+
        theme_minimal()

tidy_data %>% 
        filter(length_notes <= 1000) %>% 
        ggplot(aes(x = length_notes)) +
        labs(x = "Note length (number of characters)", 
             y = "Number of observations (log10)", 
             title = "Length of notes (length ≤ 1000)")+
        geom_histogram(binwidth = 1) +
        scale_y_log10()+
        theme_minimal()

```

```{r}
#| label: figs length notes ~ quality grade
#| code-fold: true

percentage_data <- tidy_data %>% 
        group_by(length_notes) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100)

ggplot(percentage_data, aes(x = length_notes, y = percentage)) +
        geom_col() +
        labs(x = "Length of Notes", 
             y = "RG %",
             title = "Percentage of RG ~ Length of notes") +
        theme_minimal()

percentage_data <- tidy_data %>%
        filter(length_notes <= 250) %>% 
        group_by(length_notes) %>%
        summarise(research_count = sum(quality_grade == "research"),
                  total_count = n()) %>%
        mutate(percentage = (research_count / total_count) * 100)

ggplot(percentage_data, aes(x = length_notes, y = percentage)) +
        geom_point(size = 0.5) +
        geom_smooth() +
        labs(x = "Length of Notes", 
             y = "RG %",
             title = "Percentage of RG ~ Length of notes (length ≤ 250)") +
        theme_minimal()

tidy_data %>% 
        filter (length_notes <= 500) %>% 
        ggplot(aes(x = length_notes, fill = quality_grade)) +
        geom_bar(position = "fill",
                 width = 1) +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Length of Notes", 
             y = "Proportion",
             title = "Length of notes (length ≤ 500) ~ Quality grade") +
        theme_minimal()

```

## Observation Fields

```{r}
#| label: figs presence obs fields
#| code-fold: true
#| fig-height: 6

tidy_data <- tidy_data %>%
        mutate(presence_obs_fields = ifelse(n_obs_fields == 0, "no", "yes"))

ggplot(tidy_data, aes(x = presence_obs_fields)) +
        geom_bar() +
        labs(x = "Presence of observation fields", 
             y = "Number of observations", 
             title = "Presence of observation fields")+
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal()
        

```

```{r}
#| label: figs presence obs fields ~ quality grade
#| code-fold: true

ggplot(tidy_data, aes(x = presence_obs_fields, fill=quality_grade)) +
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Presence of observation fields", 
             y = "Number of observations", 
             title = "Presence of observation fields")+
        theme_minimal()
        
```

```{r}
#| label: figs number obs fields
#| code-fold: true
#| fig-height: 6
ggplot(tidy_data,aes(x = n_obs_fields))+
        geom_bar()+
        labs(x = "Number of observation fields", 
             y = "Number of observations (log10)",
             title = "Number of observation fields")+
        scale_y_log10()+
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal()

ggplot(tidy_data,aes(x = n_obs_fields, fill= quality_grade))+
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Number of observation fields", 
             y = "Proportion", 
             title = "Number of observation fields ~ Quality grade")+
        theme_minimal()
```

## Projects

```{r}
#| label: figs presence in projects
#| code-fold: true
#| fig-height: 6
#| 
tidy_data <- tidy_data %>%
        mutate(presence_projects = ifelse(n_projects == 0, "no", "yes"))

ggplot(tidy_data, aes(x = presence_projects)) +
        geom_bar() +
        labs(x = "Presence of projects", 
             y = "Number of observations", 
             title = "Added to a traditional project")+
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal() 

```

```{r}
#| label: figs presence in projects ~ quality grade
#| code-fold: true
ggplot(tidy_data,aes(x = presence_projects, fill= quality_grade))+
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Number of projects", 
             y = "Proportion", 
             title = "Number of projects ~ Quality grade")+
        theme_minimal()  

```

```{r}
#| label: figs n projects
#| code-fold: true
#| fig-height: 6
ggplot(tidy_data,aes(x = n_projects))+
        geom_bar()+
        labs(x = "Number of projects", 
             y = "Number of observations",
             title = "Number of projects")+
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal()

ggplot(tidy_data,aes(x = n_projects, fill= quality_grade))+
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Number of projects", 
             y = "Proportion", 
             title = "Number of projects ~ Quality grade")+
        theme_minimal()

```

### Microscopy

```{r}
#| label: fig microscopy_project
#| code-fold: true
#| fig-height: 6
ggplot(tidy_data, aes(x = microscopy_project)) +
        geom_bar() +
        labs(x = "Added to \"Mushroom microscopy\" project", 
             y = "Number of observations", 
             title = " Mushroom microscopy project ~ Quality grade ")+
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal() 

```

```{r}
#| label: fig microscopy_project ~ quality grade
#| code-fold: true
ggplot(tidy_data, aes(x = microscopy_project, fill = quality_grade)) +
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Added to \"Mushroom microscopy\" project", 
             y = "Proportion", 
             title = " Mushroom microscopy project ~ Quality grade ") +
        theme_minimal()
```

## Tags

```{r}
#| label: fig presence tags
#| code-fold: true
#| fig-height: 6
tidy_data <- tidy_data %>%
        mutate(presence_tags = ifelse(n_projects == 0, "no", "yes"))

ggplot(tidy_data, aes(x = presence_tags)) +
        geom_bar() +
        labs(x = "Presence of tags", 
             y = "Number of observations", 
             title = "Presence of tags")+
        geom_text(stat = "count", aes(label = after_stat(count), vjust = -1)) +
        theme_minimal() 

ggplot(tidy_data,aes(x = presence_tags, fill= quality_grade))+
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Number of projects", 
             y = "Proportion", 
             title = "Number of projects ~ Quality grade")+
        theme_minimal() 


```

```{r}
#| label: fig number tags
#| code-fold: true

ggplot(tidy_data,aes(x = n_tags))+
        geom_bar()+
        labs(x = "Number of tags", 
             y = "Number of observations (log10)",
             title = "Number of tags")+
        scale_y_log10()+
        theme_minimal()

ggplot(tidy_data,aes(x = n_tags, fill= quality_grade))+
        geom_bar(position = "fill") +
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        labs(x = "Number of tags", 
             y = "Proportion", 
             title = "Number of projects ~ Quality grade")+
        theme_minimal()
```

## User

```{r}
#| label: fig user
#| code-fold: true
tidy_data %>% 
        mutate(user.created_at = as.Date(user.created_at)) %>% 
        mutate(user.created_at_month = floor_date(user.created_at,"month")) %>% 
        ggplot(aes(x=user.created_at_month))+
        geom_bar()+
        scale_x_date(breaks = "6 months",
                     date_labels = "%b %Y") +
        #labs(x = "t", y = "Count", title = " Uploads per day ")+
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

tidy_data %>% 
        mutate(user.created_at = as.Date(user.created_at)) %>% 
        mutate(user.created_at_month = floor_date(user.created_at,"month")) %>% 
        ggplot(aes(x=user.created_at_month,fill= quality_grade))+
        geom_bar(position = "fill")+
        scale_fill_manual(values = c(research= "#a8cc08", needs_id= "#ffee91",casual= "#aaaaaa"))+
        #labs(x = "t", y = "Count", title = " Uploads per day ")+
        theme_minimal() 


```
