---
title: "Overall model with alterations"
format: html
editor: visual
---

```{r}
#| output: false
library(tidyverse)
tidy_data <- read_csv("../data/tidy_data/tidy_data_verifiable.csv")
tidy_data <- tidy_data %>% 
        select("id", "uri", "quality_grade", "created_at", "taxon_id", "n_photos",
               "microscopy_project","days_since_upload", "cat_notes", "cat_obs_fields",
               "cat_tags","cat_projects", "created_at_year", "observed_on_year",
               "created_at_month","observed_on_month", "country","kingdom", "phylum", "class",
               "order", "family","genus","taxon_rank", "sum_pixels", "n_obs_fields", 
               "n_tags", "n_projects", "length_notes")
```

```{r}

data <- tidy_data %>% 
        filter(!is.na(sum_pixels) & n_photos != 0) %>% 
        mutate(mean_pixels = sum_pixels/n_photos,
               log_photos = log10(n_photos),
               log_sp = log10(sum_pixels),
               log_mp = log10(mean_pixels),
               rg=as.numeric(quality_grade == "research"),
               log_dsu = log10(days_since_upload + 1),
               log_notes = log10(length_notes + .Machine$double.eps),
               log_obs_fields = log10(n_obs_fields + .Machine$double.eps),
               log_tags = log10(n_tags + .Machine$double.eps),
               log_projects = log10(n_projects + .Machine$double.eps),
               )

```

Remove phylum categories that have less than 10 observations:

```{r}
data <- data %>%
        filter(phylum != "Not identified") %>% 
        group_by(phylum) %>% 
        filter(n() >= 10) %>% 
        ungroup 
```

Initial idea was to add in date of observation and taxa as a random effects, since these are factors out of our control but they can influence probability of reaching RG:

```{r}
fit_1 <- lme4::glmer(rg ~ log_dsu + log_photos + log_notes + log_tags + log_obs_fields + log_mp +  microscopy_project + log_photos:microscopy_project + (1 | created_at_month) + (1 | phylum), data= data, family = binomial,  na.action = na.fail)


summary(fit_1)
```

```{r}
dredge_results_1 <- MuMIn::dredge(fit_1, rank="AIC")
```

Model 1 runs, but gets warnings. AIC gets stuck. Seems like convergence issues.

Alternative: keep only phylum as random and month as fixed:

```{r}
fit_phylum_random <- lme4::glmer(rg ~ log_dsu + log_photos + log_notes + log_tags + log_obs_fields + microscopy_project + log_photos:microscopy_project + created_at_month + (1 | phylum), data= data, family = binomial,  na.action = na.fail)

summary(fit_phylum_random)
```

```{r}
dredge_phylum_random <- MuMIn::dredge(fit_phylum_random, rank="AIC")
```

Same issue as Model 1.

Alternative: keep only month as random, and phylum as fixed:

```{r}
fit_month_random <- glm(rg ~ log_dsu + log_photos + log_notes + log_tags + log_obs_fields + log_mp + microscopy_project + log_photos:microscopy_project + phylum + (1|created_at_month), data= data, family = binomial,  na.action = na.fail)

summary(fit_month_random)
```

```{r}
dredge_month_random <- MuMIn::dredge(fit_month_random, rank="AIC")
```

Same issue.

Alternative: keep everything as fixed effects.

```{r}
fit_fixed_only <- glm(rg ~ log_dsu + log_photos + log_notes + log_tags + log_obs_fields + log_mp + microscopy_project + log_photos:microscopy_project + created_at_month + phylum, data= data, family = binomial,  na.action = na.fail)

summary(fit_fixed_only)

```

```{r}
dredge_fixed_only <- MuMIn::dredge(fixed_only, rank="AIC")
```

I forgot log_projects!!! Running model with fixed parameters only that work with log_projects:

```{r}

fit_fixed_only_projects <- glm(rg ~ log_dsu + log_photos + log_notes + log_tags + log_obs_fields + log_projects + log_mp + microscopy_project + log_photos:microscopy_project + created_at_month + phylum, data= data, family = binomial,  na.action = na.fail)

summary(fit_fixed_only_projects)

```

```{r}
dredge_fixed_only_projects <- MuMIn::dredge(fit_fixed_only_projects, rank="AIC")
```

Results were essentialy the same. AIC once again selected full model + full model without interaction as the best ones

#### Is the issue the machine eps? Trying out log+1 instead just to check

Models run but with warning, AIC gets stuck.

```{r}
data_log_change <- tidy_data %>% 
        filter(!is.na(sum_pixels) & n_photos != 0) %>% 
        mutate(mean_pixels = sum_pixels/n_photos,
               log_photos = log10(n_photos),
               log_sp = log10(sum_pixels),
               log_mp = log10(mean_pixels),
               rg=as.numeric(quality_grade == "research"),
               log_dsu = log10(days_since_upload + 1),
               log_notes = log10(length_notes + 1),
               log_obs_fields = log10(n_obs_fields + 1),
               log_tags = log10(n_tags + 1),
               log_projects = log10(n_projects + 1),
               )

```

```{r}
fit_log_change <- lme4::glmer(rg ~ log_dsu + log_photos + log_notes + log_tags + log_obs_fields + log_mp +  microscopy_project + log_photos:microscopy_project + (1 | created_at_month) + (1 | phylum), data= data_log_change, family = binomial,  na.action = na.fail)


```

```{r}
summary(fit_log_change)
```

AIC doesn't work:

```{r}
dredge_log_change <- MuMIn::dredge(fit_log_change, rank="AIC")
```

Try change in logs, but with month as fixed effect:

```{r}
fit_log_change_2 <- lme4::glmer(rg ~ log_dsu + log_photos + log_notes + log_tags + log_obs_fields + log_mp +  microscopy_project + log_photos:microscopy_project + created_at_month + (1 | phylum), data= data_log_change, family = binomial,  na.action = na.fail)

```

```{r}
summary(fit_log_change_2)
```

```{r}
dredge_log_change_2 <- MuMIn::dredge(fit_log_change_2, rank="AIC")
```

Try change in logs, but with phylum as fixed effect:

```{r}
fit_log_change_3 <- lme4::glmer(rg ~ log_dsu + log_photos + log_notes + log_tags + log_obs_fields + log_mp +  microscopy_project + log_photos:microscopy_project + (1 |created_at_month) + phylum, data= data_log_change, family = binomial,  na.action = na.fail)
```

```{r}
dredge_log_change_3 <- MuMIn::dredge(fit_log_change_3, rank="AIC")
```

#### Is it a log problem? Changing back to categories to check

Answer: Models run without any warnings but once again AIC runs never ending warnings.

```{r}
fit_cat_change <- lme4::glmer(rg ~ log_dsu + log_photos + cat_notes + cat_tags + cat_obs_fields + log_mp +  microscopy_project + log_photos:microscopy_project + (1 | created_at_month) + (1 | phylum), data= data, family = binomial,  na.action = na.fail)

summary(fit_cat_change)
```

```{r}
fit_cat_change <- MuMIn::dredge(fit_cat_change, rank="AIC")
```

```{r}
fit_cat_change_2 <- lme4::glmer(rg ~ log_dsu + log_photos + cat_notes + cat_tags + cat_obs_fields + log_mp +  microscopy_project + log_photos:microscopy_project + created_at_month + (1 | phylum), data= data, family = binomial,  na.action = na.fail)

summary(fit_cat_change_2)
```

```{r}
fit_cat_change_2 <- MuMIn::dredge(fit_cat_change_2, rank="AIC")
```

```{r}
fit_cat_change_3 <- lme4::glmer(rg ~ log_dsu + log_photos + cat_notes + cat_tags + cat_obs_fields + log_mp +  microscopy_project + log_photos:microscopy_project + (1 | created_at_month) +  phylum, data= data, family = binomial,  na.action = na.fail)

summary(fit_cat_change_3)
```

```{r}
fit_cat_change_3 <- MuMIn::dredge(fit_cat_change_3, rank="AIC")
```
