---
title: "Ordered tables of community variables"
format: html
editor: visual
---

```{r}
#| code-fold: true
#| output: false
library(tidyverse)
library(knitr)
country_obser_iders <- read_csv("../data/country_obser_iders.csv") %>% 
        filter(country != "Vatican City" & country != "Monaco")

region_obser_iders <-  read_csv("../data/region_obser_iders.csv")

country_population <- read_csv("../data/pop_data_countries.csv")  %>% 
        filter(country != "Vatican City" & country != "Monaco")

country_obser_iders <- left_join(country_obser_iders,country_population, by = "country") 

country_obser_iders <- country_obser_iders %>% 
        mutate(n_obser_normalised = n_observers / population_2023,
               prop_obsers_iders = n_observers / n_identifiers,
               n_iders_normalised = n_identifiers / population_2023)

region_fungi_prop <- read_csv("../data/region_fungi_prop.csv")

country_fungi_prop <- read_csv("../data/country_fungi_prop.csv") %>% 
        filter(country != "Vatican City" & country != "Monaco")
```

#### 1. Number of observers per country:

```{r}
#| code-fold: true
country_obser_iders %>%
  select(country, n_observers) %>%
  arrange(desc(n_observers)) %>% 
  kable()
         
 
```

#### 2. N_observes normalised to population size per country (n_observers / population_2023):

```{r}
#| code-fold: true
country_obser_iders %>%
  select(country, n_obser_normalised) %>%
  arrange(desc(n_obser_normalised)) %>% 
  kable()


```

#### 3. Number of identifiers per country

```{r}
#| code-fold: true
country_obser_iders %>%
  select(country, n_identifiers) %>%
  arrange(desc(n_identifiers)) %>% 
  kable()
```

#### Number of identifiers per country (normalised)

```{r}
#| code-fold: true
country_obser_iders %>%
  select(country, n_iders_normalised) %>%
  arrange(desc(n_iders_normalised)) %>% 
  kable()
```

#### 4. Proportion of IDers to observers

```{r}
#| code-fold: true
country_obser_iders %>%
  select(country, prop_iders_obsers) %>%
  arrange(desc(prop_iders_obsers)) %>% 
  kable()
```

#### 5. Proportion of fungi in GBIF filtered observations

```{r}
#| code-fold: true
country_fungi_prop %>%
        filter(platform == "gbif") %>%
        select(country, prop_fungi) %>%
        arrange(desc(prop_fungi)) %>% 
        kable()
```

#### iNat filtered observations

```{r}
#| code-fold: true

country_fungi_prop %>%
        filter(platform == "inat") %>%
        select(country, prop_fungi) %>%
        arrange(desc(prop_fungi)) %>% 
        kable()
```

#### 7. Number of mycological societies per country

```{r}
#| code-fold: true
#| warning: false
n_myco_orgs <- read_csv2("../data/europe_mycological_orgs.csv")

n_myco_orgs[n_myco_orgs == "no data" | n_myco_orgs == "not applicable"] <- NA

#remove countries that were in mycological societies manuscript but not in iNat polygon 
countries_remove <- setdiff( n_myco_orgs$country, country_obser_iders$country)
n_myco_orgs <- n_myco_orgs %>% 
        filter(!country %in% countries_remove)

n_myco_orgs$n_orgs <- as.numeric(n_myco_orgs$n_orgs)
n_myco_orgs$year_oldest_org <- as.numeric(n_myco_orgs$year_oldest_org)
n_myco_orgs$log_n_orgs <- as.numeric(gsub(",", ".", n_myco_orgs$log_n_orgs))

n_myco_orgs %>%
        select(country, n_orgs) %>%
        arrange(desc(n_orgs)) %>% 
        kable()
```

#### 8. Number of CC projects per country

```{r}
#| code-fold: true
#| warning: false
n_cc_proj <- read_csv2("../data/europe_cc_projects.csv")

#remove countries that were in mycological societies manuscript but not in iNat polygon 
countries_remove <- setdiff(n_cc_proj$country, country_obser_iders$country)
n_cc_proj <- n_cc_proj %>% 
        filter(!country %in% countries_remove)


n_cc_proj %>%
        select(country, n_cc_projects) %>%
        arrange(desc(n_cc_projects)) %>% 
        kable()
```

### 9. Number of main observers per country

```{r}
tidy_data <- read_csv("../data/tidy_data/tidy_data_verifiable.csv") %>% 
        filter(country != "Vatican City" & country != "Monaco") %>% 
        select(country, user_id, n_identifications, n_identifiers)
        

# user - country combinations 
user_country_counts <- tidy_data %>%
        group_by(user_id, country) %>%
        summarise(count = n(), .groups = "drop")


# Keep the most frequent country per user, allowing ties
main_country_per_user <- user_country_counts %>%
        group_by(user_id) %>%
        slice_max(count, n = 1, with_ties = TRUE) %>% #keep only main one except for ties
        mutate(tie = n() > 1) %>%  #add T/F tag to ties
        ungroup()

#count n times ties occured
countries_ties <- main_country_per_user  %>%
        filter(tie == TRUE) %>% 
        summarise(num_users_with_ties = n_distinct(user_id))

#for some users ties occur more than once 
user_tie_count <- main_country_per_user %>%
     filter(tie == TRUE) %>%  # Keep only rows with ties
     group_by(user_id) %>%   # Group by user_id
     summarise(tie_count = n())


n_main_users_per_country <- main_country_per_user %>%
  group_by(country) %>%
  summarise(user_count = n(), .groups = "drop")

n_main_users_per_country <- left_join(n_main_users_per_country,country_population, by = "country") 

n_main_users_per_country <- n_main_users_per_country %>% 
        mutate(n_main_obsers_normalised = user_count / population_2023)
```

#### 10. Average number of identifiers and identifications per observation per country

```{r}
avg_identifiers_obs<- tidy_data %>% 
        group_by(country) %>%
        summarise(avg_n_identifiers = mean(n_identifiers))


avg_identifications_obs<- tidy_data %>% 
        group_by(country) %>%
        summarise(avg_n_identifications = mean(n_identifications))
```

#### Single comparison table (missing data from edible species per country)

```{r}
#| code-fold: true

n_observers <- country_obser_iders %>%
        arrange(desc(n_observers)) %>% 
        select(country) %>%  
        rename(n_observers = country)

n_obser_normalised <- country_obser_iders %>%
        arrange(desc(n_obser_normalised)) %>% 
        select(country) %>% 
        rename(n_obser_normalised = country)

n_identifiers <- country_obser_iders %>%
        arrange(desc(n_identifiers)) %>% 
        select(country) %>%
        rename(n_identifiers = country)

n_iders_normalised <- country_obser_iders %>%
        arrange(desc(n_iders_normalised)) %>% 
        select(country) %>%
        rename(n_iders_normalised = country)

prop_iders_obsers <- country_obser_iders %>%
        arrange(desc(prop_iders_obsers)) %>% 
        select(country) %>%
        rename(prop_iders_obsers = country)

prop_fungi_gbif <- country_fungi_prop %>%
        filter(platform == "gbif") %>%
        arrange(desc(prop_fungi)) %>% 
        select(country) %>%
        rename(prop_fung_gbif = country)

prop_fungi_inat <-country_fungi_prop %>%
        filter(platform == "inat") %>%
        arrange(desc(prop_fungi)) %>% 
        select(country) %>%
        rename(prop_fung_inat = country)

n_orgs <- n_myco_orgs %>%
        arrange(desc(n_orgs)) %>% 
        select(country) %>%
        rename(n_orgs = country)

log_n_orgs <- n_myco_orgs %>%
        arrange(desc(log_n_orgs)) %>% 
        select(country) %>%
        rename(log_n_orgs = country)

n_cc_proj <- n_cc_proj %>%
        arrange(desc(n_cc_projects)) %>% 
        select(country) %>%
        rename(n_cc_proj= country)

n_main_observers <- n_main_users_per_country %>%
        arrange(desc(user_count)) %>% 
        select(country) %>%
        rename(n_main_observers= country)

n_main_observers_norm <- n_main_users_per_country %>% 
        arrange(desc(n_main_obsers_normalised)) %>% 
        select(country) %>%
        rename(n_main_observers_norm = country)


avg_identifiers_obs <- avg_identifiers_obs %>%
        arrange(desc(avg_n_identifiers)) %>% 
        select(country) %>%
        rename(avg_n_identifiers = country)

avg_identifications_obs <- avg_identifications_obs  %>%
        arrange(desc(avg_n_identifications)) %>% 
        select(country) %>%
        rename(avg_n_identifications = country)


n_observers <- n_observers$n_observers
n_obser_normalised <- n_obser_normalised$n_obser_normalised
n_identifiers <- n_identifiers$n_identifiers
n_iders_normalised <- n_iders_normalised$n_iders_normalised
prop_iders_obsers <- prop_iders_obsers$prop_iders_obsers
prop_fungi_gbif <- prop_fungi_gbif$prop_fung_gbif
prop_fungi_inat <- prop_fungi_inat$prop_fung_inat
n_orgs <- n_orgs$n_orgs
log_n_orgs <- log_n_orgs$log_n_orgs
n_cc_proj <- n_cc_proj$n_cc_proj
n_main_observers <- n_main_observers$n_main_observers
n_main_observers_norm <- n_main_observers_norm$n_main_observers_norm 
avg_identifiers_obs <- avg_identifiers_obs$avg_n_identifiers
avg_identifications_obs <- avg_identifications_obs$avg_n_identifications


length(n_orgs) <- length(n_observers)
length(log_n_orgs) <- length(n_observers)
length(n_cc_proj) <- length(n_observers)


country_orders <- data.frame(index = seq_along(n_observers), n_observers, n_obser_normalised, n_main_observers, n_main_observers_norm, n_identifiers, n_iders_normalised,
           prop_iders_obsers, prop_fungi_gbif, prop_fungi_inat, n_orgs,
           log_n_orgs, n_cc_proj, avg_identifiers_obs, avg_identifications_obs, 
           index = seq_along(n_observers))

kable(country_orders)

writexl::write_xlsx(country_orders, "country_orders.xlsx")
```
